---
title: 'EES 3310/5310 Lab #1'
author: "put your name here"
date: "Monday Jan. 24, 2022. Due Wed. Feb. 2, 2022"
output:
  html_document:
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
    includes:
      in_header: ees3310.sty
fontsize: 12pt
subtitle: Exercises on Units, Energy, and Power
---

```{r setup, include=FALSE}
# This section sets up some options on knitr for processing the RMarkdown file
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)
```
```{r initialize, include=FALSE}
# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
library(tidyverse)

script_dir = "_scripts"

source(file.path(script_dir, "format_md.R"), chdir =  TRUE)
```
# Exercises on Units, Energy, and Power

**_I have worked Exercise #1 as an example to show you how you can use 
RMarkdown to solve problems and write up the answers. 
You will solve the other exercises for the lab using the same methods._**

## Worked Example

Consider this challenge of calculating how much energy a light bulb uses in a
day.

### Part (a)

> (a) A Joule (J) is an amount of energy, and a watt (W) is a rate of using 
> energy, defined as 1 Watt = 1 Joule per second (abbreviated 1 W = 1 J/s). 
> How many Joules of energy are required to run a 100-W light bulb for one day?

In this code chunk, I define several conversion factors to make it easy to
convert between seconds, hours, and days and to convert from Watts to 
Joules per day.

```{r example_a}
seconds_per_hour = 60 * 60 # number of seconds in one hour
hours_per_day = 24 # number of seconds in one day
light_bulb_power = 100
joules_per_day = light_bulb_power * seconds_per_hour * hours_per_day

print(joules_per_day)
```

It can be hard to read a long number like that, so we can use an R function
I wrote, called `format_md` to add commas:

```{r formatting_example}
print(format_md(joules_per_day, comma = TRUE))
```

We could also format this using scientific notation:
```{r scientific_formatting_example, results="asis"}
print(format_md(joules_per_day, digits = 2, format = "scientific"))
```

You can find details about using the `format_md` function in the 
`README.md` document for this lab. 

**Answer:** A `r light_bulb_power` watt light bulb burns `r light_bulb_power` 
Joules per second, so the light bulb burns 
`r format_md(light_bulb_power * seconds_per_hour, digits = 2, format = "normal", comma = TRUE)` 
Joules per hour and
`r format_md(joules_per_day, digits = 2, format = "normal", comma = TRUE)` 
Joules per day.

### Part (b)

> (b) Burning coal yields about 3.0 &times; 10^7^J of energy per 
> kilogram of coal burned. 
> Assuming that the coal power plant is 30% efficient,
> how much coal has to be burned to light that light bulb for one day?


To solve this problem, I start by putting the numbers into R:
```{r example_b_initialize}
joules_heat_per_kg_coal = 3.0E+7 # you tell R about numbers in scientific 
                                 # notation by using `xxEyy` to indicate
                                 # xx times 10 to the yy power.
efficiency = 0.30
```
One kg of coal produces `r format_md(joules_heat_per_kg_coal, format = "scientific")`
Joules of heat. The efficiency is `r efficiency`, so
`r 100 * efficiency`% of the heat is turned into electricity.
```{r example_b_calculation}
joules_electricity_per_kg_coal = joules_heat_per_kg_coal * efficiency
```

This means that 1 kg coal produces 
`r format_md(joules_electricity_per_kg_coal, digits=2, comma=TRUE)` Joules 
of electricity.
We can turn this around to figure out how many kg of coal it takes to 
generate 1 Joule of electricity:
```{r example_b_calc_2}
kg_coal_per_joule_electricity = 1 / joules_electricity_per_kg_coal
kg_coal_per_day = joules_per_day * kg_coal_per_joule_electricity
```
**Answer:**
One Joule of electricity takes 
`r format_md(kg_coal_per_joule_electricity, digits=2, format="scientific")` 
kg of coal, and the 
`r format_md(joules_per_day, digits=2, comma=TRUE)` Joules of electricity 
required to power the light bulb for one day takes 
`r format_md(kg_coal_per_day, digits = 1, comma=TRUE)` kg of coal.



## Exercise 1

> This exercise asks you to calculate how many Joules of energy you
> can get for a dollar from different sources of energy.

### Part 1(a)

> (a) A gallon of gasoline carries with it about 1.3 &times; 10^8^ J of energy. 
> Given a price of $3 per callon, how many Joules can you get for a dollar?

_Put your own text and R code here to answer Exercise 1(a). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example above._

```{r ex_1_a}
# Put your R code here to answer the question
# 
# You may use multiiple chunks of R code, with text in between to explain
# what you are doing.
# 
# Hint: In R, 1.3E+8 is how you say 1.3 times 10 to the 8th power.
```

**Answer:** You can get ... Joules per dollar from gasoline. 
_Edit this to put a full answer, following the model of the worked example._

### Part 1(b)

> (b) Electricity goes for about $0.05 per kilowatt hour. A kilowatt hour is 
> just a different way to express energy, because a Watt is a Joule per second, 
> and a kilowatt hour is the number of Joules (the amount of energy) one would 
> get from running 1000 Watts for one hour (3,600 seconds). 
> How many Joules of electricity can you get for a dollar?

_Put your own text and R code here to answer Exercise 1(b). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example above._


```{r ex_1_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ... 
_Edit this to put a full answer, following the model of the worked example._

### Part 1(c)

> (c) A standard cubic foot of natural gas carries with it about 
> 1.1 &times; 10^6^ Joules of energy. 
> You can get about 5 &times; 10^5^ British Thermal Units (BTUs) of gas
> for a dollar, and there are about 1,030 BTUs in a standard cubic foot. 
> How many Joules of energy in the form of natural gas can you get for a dollar?

_Put your own text and R code here to answer Exercise 1(c). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example above._

```{r ex_1_c}
# Put your R code here to answer the question
```

**Answer:** put your answer here ... 
_Edit this to put a full answer, following the model of the worked example._

### Part 1(d)

> (d) A ton of coal holds about 3.2 &times; 10^10^ J of energy and costs 
> about $40.
> How many Joules of energy in the form of coal can you get for a dollar?

_Put your own text and R code here to answer Exercise 1(d). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_1_d}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

### Part 1(e)

> (e) Corn oil costs about $0.10 per fluid ounce wholesale. A fluid ounce 
> carries about 240 dietary Calories (which a scientist would call 
> kilocalories). 
> A dietary Calorie is about 4200 J. 
> How many Joules of energy in the form of corn oil can you get for a dollar?

_Put your own text and R code here to answer Exercise 1(e). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_1_e}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

### Part 1(f)

> (f) Now we compare the different energy sources.
> Rank these five energy sources from cheap to expensive. 
> What is the range of prices per Joule?

_Put your own text and R code here to answer Exercise 1(f). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_1_f}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._


## Exercise 2

> In this exercise, we compare the energy it took to produce the concrete in the
> Hoover Dam (outside Las Vegas) to the energy the dam produces from 
> hydroelectric generation. 

### Part 2(a)

> The Hoover Dam produces 2 &times; 10^9^ W of electricity. 
> It is composed of 7 &times; 10^9^ kg of concrete.
> It requires 1 MJ of energy (1 megajoule, 1,000,000 Joules) to produce 
> each kilogram of concrete. 
> How much energy did it take to produce the concrete for the dam? 

_Put your own text and R code here to answer Exercise 2a. Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_3_a}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

### Part 2(b)

> How long is the payback time for the dam to generate as much energy in 
> electricity as it took to produce the concrete?

_Put your own text and R code here to answer Exercise 2b. Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_2_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

## Exercise 3

> It takes approximately 2.0 &times; 10^9^ J of energy 
> to manufacture 1 m^2^ of crystalline-silicon photovoltaic cell.
> An average of 250 W/m^2^ falls on the Earth. 
> Assume that the solar cell is 18% efficient (that is, it converts 18% of
> the energy from sunlight into electricity). 
> Calculate how long it would take for the solar cell to repay the 
> energy it took to manufacture it.

_Put your own text and R code here to answer Exercise 3. Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_3}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

## Exercise 4 (Grad students only)

Only graduate students are required to do this exercise.

> This is one of those job-interview questions to see how creative you are, 
> analogous to, "How many airplanes are over Chicago at any given time?" 
> You need to make stuff up to get a rough estimate and demonstrate your 
> ability to think creatively. 
> The question is: 
> _What is the efficiency of energy production from growing corn?_

### Part 4(a)
 
> (a) Assume that sunlight deposits 250 W/m^2^ of energy on a corn field, 
> averaging over the day-night cycle. 
> There are approximately 4,200 J per dietary Calorie. 
> How many Calories of energy are deposited on a square
> meter of field over the growing season? 
>
> (Note: the word "calorie" has two different meanings. Physicists and chemists, 
> use "calorie" (with a lower-case "c") to refer to a thermodynamic unit of heat, 
> but nutritionists use the word Calorie (with a capital 'C')
> to mean 1 kilocalorie (1000 thermodynamic calories). When you see "Calories" 
> on a food label, it means kilocalories. 
> To keep this exercise simple, I have written it so you only need to think in 
> terms of dietary Calories.)

_Put your own text and R code here to answer Exercise 4(a). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example above._

```{r ex_4_a}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._

### Part 4(b)
 
> (b) Now guess how many ears of corn grow per square
> meter, and guess what the number of dietary Calories is that you get for eating 
> an ear of corn. Compare the sunlight energy with the corn energy to get the 
> efficiency.

_Put your own text and R code here to answer Exercise 4(a). Do the calculation
using R and also explain what you are doing with your text, following the 
model of the worked example._

```{r ex_4_b}
# Put your R code here to answer the question
```

**Answer:** put your answer here ...
_Edit this to put a full answer, following the model of the worked example._
