---
title: "Assignment for EES 3310/5310 Lab #1"
date: "January 24, 2022"
output: 
  github_document: default
---
```{r setup, include=FALSE}
# This section sets up some options on knitr for processing the RMarkdown file
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)
```
```{r initialize, include=FALSE}
# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
library(tidyverse)

script_dir = "_scripts"

source(file.path(script_dir, "format_md.R"), chdir =  TRUE)
```

# Assignment for EES 3310/5310 Lab #1


<!-- badges: start -->
[![License: CC BY 4.0](https://img.shields.io/badge/License-CC%20BY%204.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)
<!-- badges: end -->

The file `lab-01-report.Rmd` is a template for the exercises in Lab #1 to use
RMarkdown.

The lab is due by 11:59 pm on Wednesday, Feb. 2.
**For this first lab assignment**, you will get full credit will be given for 
trying, even if your R code doesn't work or if you get the answers wrong. 

**Starting next week**, labs will also be graded for correctness and for 
effective use of RMarkdown in writing up and explaining your answers.

In the report, I give a worked example as a model of how to answer problems
using RMarkdown.

There are three exercises for undergraduates and four for graduate students.

## Instructions:

* Accept the assignment in GitHub Classroom.
* On your web browser, go to the repository for the assignment.
* In RStudio, create a new project, choose "Version Control". Then choose 
  "Git" and paste the URL for your assignment repository.
* In your RStudio project, use the template "`lab-01-report.Rmd`" and 
  fill in text and R code to answer the questions. 
  There is a worked example to help you understand what you need to do.
* When you are done, knit your `.Rmd` file into a `.pdf` file or a 
  Word (`.docx`) file.
* Use git to commit your changes (including the edits to `lab-01-report.Rmd`
  and the new `.pdf` or `.docx` file) to your local git repository.
  You can do this from the "Git" tab in RStudio.
* Push the changes from your local git repository to GitHub. You can do this 
  using the "Push" button (the green up arrow) in the "Git" tab in RStudio.
* When you are finished with the lab, or have gotten as far as you can go,
  go to the Brightspace discussion board and post a comment about your
  experience using these tools (R, RStudio, and Git) to do these exercises.
  
  Comment about what was clear, what was unclear or difficult, and any ideas 
  you have about what would make it easier to do these kinds of labs in the 
  future.

As you work on these projects, I advise using git to commit changes frequently 
as you work and push those commits to GitHub.

## Formatting Numbers in R

When you use RMarkdown to write reports, it is helpful to be able to 
format numbers effectively. For this lab and future ones, I have supplied you
with a convenient function called `format_md`, which allows you to take a 
number in R and format it flexibly to look good in your report.

**There are many ways to format numbers in R and using the `format_md` 
function is completely optional.** It's there if you find it helpful, but 
you don't have to use it if you don't want to, or if you prefer to use
a different way to format numbers.

If you have a variable `x` that contains a number, you can format it 
using `format_md(x)`, and you can specify additional arguments to 
`format_md` to control exactly how the number gets formatted.

Let's start by assigning the value of &pi; to `x`:
```{r include=TRUE}
x = pi
format_md(x)
```

Now, suppose we only want to format the number to only have two additional 
digits after the initial 3:
```{r include=TRUE}
format_md(x, digits = 2)
```
Here is another example:
```{r include=TRUE}
format_md(x * 10, digits = 2)
```
```{r include=TRUE}
format_md(x * 100, digits = 4)
```
```{r include=TRUE}
format_md(x, digits = 10)
```

We can also format the number using scientific notation:
```{r include=TRUE}
y = pi * 1000000
format_md(y, digits = 2, format = "scientific")
```
Or we can format long numbers to put commas separating the thousands, 
millions, etc:
```{r include=TRUE}
format_md(y, comma = TRUE)
```

### Different ways to format numbers

If you want to express a number in scientific notation, you can do this, 
using `format_md`. In the example below `1E6` is a convenient way to
write 10^6^ in R. You can also write `5E10` to write 5 &times; 10^10^,
and so forth.
```{r scientific_formatting_example, results="asis"}
x = pi * 1E6
print(format_md(x, digits = 2, format = "scientific"))
```

When you are writing Markdown text, there are three good ways to express a
number in scientific notation: If we want to express $1.23 \times 10^{7}$, we
can do this as
```
1.23 &times; 10^7^
```
(note the `^` both before and after the exponent and the `;` in `&times;`) 
or 
```
$1.23 \times 10^{7}$
```
(note the dollar signs before and after, and the curly braces `{}` around the 
exponent)
or 
```{r echo=TRUE, eval=FALSE}
`r format_md(1.23E7, digits=2, format = "scientific")`
```
which produces `r format_md(1.23E7, digits=2, format = "scientific")`.
Note the backward quotes `` ` `` before and after the R expression and 
the `r` after the first `` ` ``, which tells RMarkdown that this is R code it 
should evaluate; also note that we used the notation `1.23E7` for the 
number $1.23 \times 10^{7}$ in R.

We can also tell R to put the numbers directly into our text so we don't have 
to type them. If we assign a value to the variable `x` like this
```{r}
x = 1.23456E7
```
then later in our text, we can put a formatted number into the middle of
our text.

This RMarkdown text
```{text}
The value of _x_ is `r format_md(x, digits = 2, format = "scientific")` and
the value of pi is `r format_md(pi, digits = 2)`.
```
becomes the following:

> The value of _x_ is `r format_md(x, digits = 2, format = "scientific")` and
> the value of pi is `r format_md(pi, digits = 2)`.
